The function--sheaf dictionary for smooth commutative group schemes over finite fields, with applications over local fields

### 2014-08-22 dictionary.tex
Going to lunch. Back soon  . . .

As indicated above, our interest in the function--sheaf dictionary for smooth commutative group schemes over finite fields comes from an application to $p$-adic representation theory. 
We found, however, that this application applies to all local fields $K$ with finite residue field $\Fq$ and all connected commutative algebraic groups $X_K$ over $K$ that do not contain a subgroup of type $\mathbb{G}_\text{a}$, such as Abelian varieties, unipotent $K$-wound groups, and of course, algebraic tori. 
Such algebraic groups $X_K$ always admit a Neron model $X$ over $R$, the ring of integers of $K$.
In this paper we show that quasicharacters of $X_K$ are geometrized and categorified by quasicharacter sheaves on the
Greenberg transform $\Gr_R(X)$ of the N\'eron model $X$; although not smooth, $\Gr_R(X)$ is a commutative group scheme over $\Fq$ and it is a projective limit of smooth commutative group schemes $\Gr^R_n(X)$.
This allows us to adapt our work on quasicharacter sheaves to construct a category $\QC(\Gr_R(X))$ of quasicharacter sheaves on $\Gr(X)$.
The ability to generalize the function--sheaf dictionary to non-connected group schemes, of which the component group of the special fibre of the Néron model $X$ is an important example, plays a crucial role in this application
since these group schemes are not connected in general.

@@

\subsection{Structure of the paper}
\todo{Describe the overall structure of the paper in enough detail that one can follow the argument; include snake lemma diagram}

We close the introduction with an overview of the structure of this paper.
After setting some basic notation in Section~\ref{sec:notation},
in Section~\ref{sec:category} we define the category of quasicharacter sheaves.
Section~\ref{sec:Frob} introduces the trace of Frobenius homomorphism $\TrFrob{G} : \QCiso{G} \to \Hom(G(k), \EEx)$.
We also define pullbacks of quasicharacter sheaves along morphisms of group schemes.  Applying pullbacks to the component group
sequence for $G$ yields the diagram
 \begin{equation} \label{eq:intro-snake}
  \begin{tikzcd}[row sep=20]
    0 \rar & \QCiso{\pi_0(G)} \rar \dar{\TrFrob{\pi_0(G)}}
    & \QCiso{G} \rar \dar{\TrFrob{G}} & \QCiso{G^0} \rar \dar{\TrFrob{G^0}} & 0\\
    0 \rar & \pi_0(G)(\Fq)^* \rar & G(\Fq)^* \rar & G^0(\Fq)^* \rar & 0,\\
  \end{tikzcd}
  \end{equation}
which is central to the proof of Theorem~\ref{thm:snake}.  Much of Sections~\ref{sec:etale}, \ref{sec:disc-isog} and \ref{sec:main}
are devoted to proving that the rows of \eqref{eq:intro-snake} are exact and determining the kernels and cokernels of the vertical maps.

@@

\todo{Comments about Joseph Bernstein's comments?}

@@

\begin{definition}\label{def:QC}
A \emph{quasicharacter sheaf on $G$} is a triple
$\qcs{L}\ceq (\gqcs{L},\mu,\phi)$ where:
\begin{enumerate}
\labitem{(QC.1)}{QC.1} $\gqcs{L}$ is a rank~one $\ell$-adic local system on $\bG$, by which we mean
a constructible $\ell$-adic sheaf on $\bG$, locally constant on each connected
component of $\bG$, whose stalks are one-dimensional $\EE$-vector spaces;
\labitem{(QC.2)}{QC.2} $\mu: \bm^* \gqcs{L} \to \LxL$ is an isomorphism of
sheaves on $\bGxG$ such that the following diagram commutes,
  where $m_3 \ceq m\circ (m\tight{1}{\times}{2}\id) = m\circ (\id\tight{2}{\times}{1} m)$;
  \[
  \begin{tikzcd}[row sep=30]
  \bm_3^*\gqcs{L} \arrow{rr}{(\bm \tight{1}{\times}{2} \id)^*\mu} \arrow[swap]{d}{(\id \tight{2}{\times}{1} \bm)^*\mu}
    &&  \bm^*\gqcs{L} \boxtimes \gqcs{L} \dar{\mu \tight{0}{\boxtimes}{1} \id} \\
    \gqcs{L} \boxtimes \bm^* \gqcs{L} \arrow{rr}{\id \boxtimes \mu}
    &&  \gqcs{L} \tight{0}{\boxtimes}{0} \LxL
  \end{tikzcd}
  \]
\labitem{(QC.3)}{QC.3} $\phi : \Frob{G}^* \gqcs{L} \to \gqcs{L}$ is an
  isomorphism of constructible $\ell$-adic sheaves on $\bG$ compatible with
  $\mu$ in the sense that the following diagram commutes.
  \[
  \begin{tikzcd}[row sep=20]
  \Frob{\GxxG}^* \bm^* \gqcs{L} \arrow{rr}{\Frob{\GxxG}^*\mu}
    && \Frob{\GxxG}^*(\LxL)\\
    \arrow[equal]{u} \bm^*  \Frob{G}^* \gqcs{L} \arrow[swap]{d}{\bm^* \phi}
    && \Frob{G}^*\gqcs{L}\boxtimes \Frob{G}^*\gqcs{L} \dar{\phi\boxtimes \phi} \arrow[equal]{u} \\
    \bm^*\gqcs{L} \arrow{rr}{\mu}
    && \LxL
  \end{tikzcd}
  \]
\end{enumerate}
\end{definition}

@@

\begin{proof}
  Let $\qcs{L}$ be a quasicharacter sheaf on $G$. We start by showing that
  $\bf^*\gqcs{L}$ is a local system of rank~one. Let $c_j :\bH^j \hookrightarrow \bH$ be any
  connected component and let $d_i : \bG^i \hookrightarrow \bG$ be the image of $j$ under the group
  homomorphism $\pi_0(\bf) : \pi_0(\bH) \to \pi_0(\bG)$.  Write $\bf^j : \bH^j \to \bG^i$
  for the restriction of $f$ to $\bH^j$.  Since $\gqcs{L}$ is locally constant on the connected
  components of $\bG$, there is a finite \'etale covering
  $a_i : X_i \to \bG^i$ such that $a_i^* (\gqcs{L}\vert_{\bG^i})$
  is constant.  Define $Y_j$ in the following diagram through pullback:
  \[
  \begin{tikzcd}[row sep=20, column sep=30]
   Y_j \dar[dashed]{g_j} \rar[dashed]{b_j} & \bH^j \dar{\bf^j} \rar[hook]{c_j} & \bH \dar{\bf} \\
   X_i \rar{a_i} & \bG^i \rar[hook]{d_i} & \bG
  \end{tikzcd}
  \]
  Then $b_j$ is a finite \'etale covering
  of $\bH^j$ and
  \[
  b_j^* \left( (\bf^* \gqcs{L})\vert_{\bH^j}\right)
  = (\bf\circ c_j \circ b_j)^*\gqcs{L}
  = (d_i\circ a_i\circ g_j)^*\gqcs{L} = g_j^* a_i^* (\gqcs{L}\vert_{\bG^i}).
  \]
  Since $a_i^* (\gqcs{L}\vert_{\bG^i})$ is a constant sheaf, its pullback
  $b_j^* \left( (\bf^*\gqcs{L})\vert_{\bH^j}\right)$ along $g_j$ is also constant.
  Thus $\bf^*\gqcs{L}$ is locally constant on
  $\bH^j$. To complete the proof that $\bf^*\gqcs{L}$ satisfies
  \ref{QC.1}
  simply observe that $\bf^*\gqcs{L}$ has rank~one since
  $(\bf^*\gqcs{L})_{\bg} = \gqcs{L}_{f(\bg)}$, for every geometric point ${\bg}$ on $\bG$.

  To see that $(\bfxf)^* \mu$ satisfies \ref{QC.2},
  apply the functor $(\bfxf)^*$
  to \ref{QC.2} for $\qcs{L}$ and use the canonical isomorphism
  $(\bfxf)^*(\LxL) \iso \bf^*\gqcs{L} \tight{-3}{\boxtimes}{-3} \bf^*\gqcs{L}$.
  To show that $f^*\qcs{L}$ satisfies
  \ref{QC.3}, apply the same functor to \ref{QC.3} for $\qcs{L}$.
  Since $f$ is a morphism of group schemes defined over $\Fq$
  it provides isomorphisms $(\bfxf)^*\Frob{\GxxG}^* \iso \Frob{\GxxG}^* (\bfxf)^*$
  and $(\bfxf)^* \bm^*\iso \bm^* \bf^*$ between functors of constructible sheaves.

  Applying $\bf^*$ and $\bf^* \tight{1}{\times}{1}\bf^*$ to \ref{QC.4} defines the action
  of $f^*$ on morphisms of quasicharacter sheaves; arguing as above shows that $f^*$ is
  a functor from $\QC(G)$ to $\QC(H)$.  Since tensor products commute with pullback in schemes,
  $f^* : \QC(G) \to \QC(H)$ is a monoidal functor.
  The diagram relating $f^* : \QC(G) \to \QC(H)$, $f^* : G(k)^* \to H(k)^*$ and Trace of Frobenius
  commutes by \cite{laumon:87a}*{1.1.1.2}, where the ambient
  hypothesis that $X$ is of finite type over $\Fq$ can be replaced by
  the hypothesis that $X$ is locally of finite type over $\Fq$.

@@

\begin{proof}
Suppose $\rho \in \Hom(\qcs{L},\qcs{L}')$.
If $\rho_0 = 0$ then the second diagram in \ref{qc.4} for $y=0$ implies that $\rho_x = 0$ for all $x\in G$, in which case $\rho$ is trivial.
On the other hand, if $\rho_0$ is nonzero then the second diagram in \ref{qc.4} for $y = -x$ implies that $\rho_x$ is nonzero for all $x$ and thus an isomorphism since the stalk of quasicharacter sheaves are one dimensional; in this case $\rho$ is an isomorphism.

>>>>

As indicated above, our interest in the function--sheaf dictionary for smooth commutative group schemes over finite fields comes from an application to $p$-adic representation theory. 
We found, however, that this application applies to all local fields $K$ with finite residue field $\Fq$ and all connected commutative algebraic groups $X_K$ over $K$ that do not contain a subgroup of type $\mathbb{G}_\text{a}$, such as Abelian varieties, unipotent $K$-wound groups, and of course, algebraic tori. 
Such algebraic groups $X_K$ always admit a Neron model $X$ over $R$, the ring of integers of $K$.
In this paper we show that quasicharacters of $X_K$ are geometrized and categorified by quasicharacter sheaves on the
Greenberg transform $\Gr_R(X)$ of the N\'eron model $X$; although not locally of finite type, $\Gr_R(X)$ is a commutative group scheme over $\Fq$ and also a projective limit of smooth commutative group schemes $\Gr^R_n(X)$.
This allows us to adapt our work on quasicharacter sheaves to construct a category $\QC(\Gr_R(X))$ of quasicharacter sheaves on $\Gr(X)$.
The ability to generalize the function--sheaf dictionary to non-connected group schemes, of which the component group of the special fibre of the Néron model $X$ is an important example, plays a crucial role in this application
since these group schemes are not connected in general.

@@

\subsection{Structure of the paper}

We close the introduction with an overview of the structure of this paper.
After setting some basic notation in Section~\ref{sec:notation},
in Section~\ref{sec:category} we define the category of quasicharacter sheaves.
Section~\ref{sec:Frob} introduces the trace of Frobenius homomorphism $\TrFrob{G} : \QCiso{G} \to \Hom(G(\FF_q), \EEx)$.
We also define pullbacks of quasicharacter sheaves along morphisms of group schemes.  Applying pullbacks to the component group
sequence for $G$ yields the diagram
 \begin{equation} \label{eq:intro-snake}
  \begin{tikzcd}[row sep=20]
    0 \rar & \QCiso{\pi_0(G)} \rar \dar{\TrFrob{\pi_0(G)}}
    & \QCiso{G} \rar \dar{\TrFrob{G}} & \QCiso{G^0} \rar \dar{\TrFrob{G^0}} & 0\\
    0 \rar & \pi_0(G)(\FF_q)^* \rar & G(\Fq)^* \rar & G^0(\FF_q)^* \rar & 0,
  \end{tikzcd}
  \end{equation}
which is central to the proof of Theorem~\ref{thm:snake}.  Much of Sections~\ref{sec:etale}, \ref{sec:disc-isog} and \ref{sec:main}
are devoted to proving that the rows of \eqref{eq:intro-snake} are exact and determining the kernels and cokernels of the vertical maps.

@@

@@

\begin{definition}\label{def:QC}
A \emph{quasicharacter sheaf on $G$} is a triple
$\qcs{L}\ceq (\gqcs{L},\mu,\phi)$ where:
\begin{enumerate}
\labitem{(QC.1)}{QC.1} $\gqcs{L}$ is a rank-one $\ell$-adic local system on $\bG$, by which we mean
a constructible $\ell$-adic sheaf on $\bG$, locally constant on each connected
component of $\bG$, whose stalks are one-dimensional $\EE$-vector spaces;
\labitem{(QC.2)}{QC.2} $\mu: \bm^* \gqcs{L} \to \LxL$ is an isomorphism of
sheaves on $\bGxG$ such that the following diagram commutes,
  where $m_3 \ceq m\circ (m\tight{1}{\times}{2}\id) = m\circ (\id\tight{2}{\times}{1} m)$;
  \[
  \begin{tikzcd}[row sep=30]
  \bm_3^*\gqcs{L} \arrow{rr}{(\bm \tight{1}{\times}{2} \id)^*\mu} \arrow[swap]{d}{(\id \tight{2}{\times}{1} \bm)^*\mu}
    &&  \bm^*\gqcs{L} \boxtimes \gqcs{L} \dar{\mu \tight{0}{\boxtimes}{1} \id} \\
    \gqcs{L} \boxtimes \bm^* \gqcs{L} \arrow{rr}{\id \boxtimes \mu}
    &&  \gqcs{L} \tight{0}{\boxtimes}{0} \LxL
  \end{tikzcd}
  \]
\labitem{(QC.3)}{QC.3} $\phi : \Frob{G}^* \gqcs{L} \to \gqcs{L}$ is an
  isomorphism of constructible $\ell$-adic sheaves on $\bG$ compatible with
  $\mu$ in the sense that the following diagram commutes.
  \[
  \begin{tikzcd}[row sep=20]
  \Frob{\GxxG}^* \bm^* \gqcs{L} \arrow{rr}{\Frob{\GxxG}^*\mu}
    && \Frob{\GxxG}^*(\LxL)\\
    \arrow[equal]{u} \bm^*  \Frob{G}^* \gqcs{L} \arrow[swap]{d}{\bm^* \phi}
    && \Frob{G}^*\gqcs{L}\boxtimes \Frob{G}^*\gqcs{L} \dar{\phi\boxtimes \phi} \arrow[equal]{u} \\
    \bm^*\gqcs{L} \arrow{rr}{\mu}
    && \LxL
  \end{tikzcd}
  \]
\end{enumerate}
\end{definition}

@@

\begin{proof}
  Let $\qcs{L}$ be a quasicharacter sheaf on $G$. It is clear \todo{Reference} that
  $\bf^*\gqcs{L}$ is a local system of rank~one, so  \ref{QC.1}  is satisfied.
  To see that $(\bfxf)^* \mu$ satisfies \ref{QC.2},
  apply the functor $(\bfxf)^*$
  to \ref{QC.2} for $\qcs{L}$ and use the canonical isomorphism
  $(\bfxf)^*(\LxL) \iso \bf^*\gqcs{L} \tight{-3}{\boxtimes}{-3} \bf^*\gqcs{L}$.
  To show that $f^*\qcs{L}$ satisfies
  \ref{QC.3}, apply the same functor to \ref{QC.3} for $\qcs{L}$.
  Since $f$ is a morphism of group schemes defined over $\Fq$
  it provides isomorphisms $(\bfxf)^*\Frob{\GxxG}^* \iso \Frob{\GxxG}^* (\bfxf)^*$
  and $(\bfxf)^* \bm^*\iso \bm^* \bf^*$ between functors of constructible sheaves.

  Applying $\bf^*$ and $\bf^* \tight{1}{\times}{1}\bf^*$ to \ref{QC.4} defines the action
  of $f^*$ on morphisms of quasicharacter sheaves; arguing as above shows that $f^*$ is
  a functor from $\QC(G)$ to $\QC(H)$.  Since tensor products commute with pullback in schemes,
  $f^* : \QC(G) \to \QC(H)$ is a monoidal functor.
  The diagram relating $f^* : \QC(G) \to \QC(H)$, $f^* : G(k)^* \to H(k)^*$ and trace of Frobenius
  commutes by \cite{laumon:87a}*{1.1.1.2}, where the ambient
  hypothesis that $X$ is of finite type over $\Fq$ can be replaced by
  the hypothesis that $X$ is locally of finite type over $\Fq$.

@@

\begin{proof}
Suppose $\rho \in \Hom(\qcs{L},\qcs{L}')$.
If $\rho_0 = 0$ then the second diagram in \ref{qc.4} for $y=0$ implies that $\rho_x = 0$ for all $x\in G$, in which case $\rho$ is trivial.
On the other hand, if $\rho_0$ is nonzero then the second diagram in \ref{qc.4} for $y = -x$ implies that $\rho_x$ is nonzero for all $x$ and thus an isomorphism since the stalk of quasicharacter sheaves are one-dimensional; in this case $\rho$ is an isomorphism.
